---
title: "Risk and Resilience in Deep Time"
subtitle: "Analysis Scripts for Paper Published in _Holocene_"
author: "C Michael Barton, Arizona State University"
date: "2023-03-27"
output: html_notebook
---

Set up libraries
```{r setup, include=FALSE, echo=T}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rcarbon)
library(scales)
library(parallel)

ncores <- detectCores()
options(Ncpus = ncores)
```


Load data
```{r load data, echo=T}
c14dates <- read_csv("c14dates.csv")

Hominin_demography <- read_csv("Hominin_demography.csv", 
                               col_types = cols(...1 = col_integer(), 
                                                home.range.radius = col_integer(), 
                                                population = col_integer()))
Hominin_demography <- Hominin_demography %>% 
  mutate(phenotype = recode(phenotype, 
                            "1 MM"="MM", 
                            "2 M-type"="M-type", 
                            "3 hybrid"="MN", 
                            "4 N-type"="N-type", 
                            "5 NN"="NN"), 
         Fitness = recode(Fitness, 
                          "hybrid higher mortality"="MN higher mortality", 
                          "hybrid lower fertility"="MN lower fertility", 
                          "hybrid lower mortality"="MN lower mortality")) %>% 
  mutate(phenotype = factor(phenotype, levels = c("MM", "M-type", "MN", "N-type", "NN")))

Hominin_demography_trajectory <- read_csv("Hominin_demography_trajectories.csv")
Hominin_demography_trajectory <- Hominin_demography_trajectory %>% 
  select(M.type.x, ends_with(".y")) %>% 
  rename(step = M.type.x, 
         "M-Type" = M.type.y, 
         "N-Type" = N.type.y, 
         MN = hybrid.y,
         MM = MM.y, 
         NN = NN.y) %>% 
  pivot_longer(2:6, names_to = "phenotype", values_to = "population") %>% 
  mutate(phenotype = factor(phenotype, levels = c("MM", "M-Type", "MN", "N-Type", "NN")))

icecores <- read_csv("icecores.csv")

rr.papers <- read_csv("rr_papers.csv")

```

### Figure 1: Journal Publications

```{r bibliometrics, echo=T, fig.width=10, fig.height=5}
rr.papers %>% 
  ggplot(aes(x=years)) + 
  geom_path(aes(y=JAS.risk/JAS.papers, color='Journal of Archaeological Science', lty='risk'), group=1, lwd=1.5) + 
  geom_path(aes(y=JAS.resilience/JAS.papers, color='Journal of Archaeological Science', lty='resilience'), group=1, lwd=1.5) + 
  geom_path(aes(y=Antiquity.risk/Antiquity.papers, color='Antiquity', lty='risk'), group=1, lwd=1.5) + 
  geom_path(aes(y=Antiquity.resilience/Antiquity.papers, color='Antiquity', lty='resilience'), group=1, lwd=1.5) + 
  scale_y_continuous(labels = scales::percent) + 
  scale_color_manual(values = c('Journal of Archaeological Science' = 'red', 'Antiquity' = 'blue')) + 
  scale_linetype_manual(values = c('risk' = 'solid', 'resilience' = 'dashed')) + 
  labs(title = "Papers Mentioning Risk or Resilience", 
       subtitle = "Change in Past 40 Years", 
       x="5-year intervals", 
       y="% of papers published", 
       color='journal:', 
       linetype='topic:') + 
  theme_bw(base_size = 30) + 
  theme(legend.position = 'bottom', legend.key.width =  unit(50, "points"))
```


## Modeling Biocultural Interaction

### Figure 2: NGRIP2 and GISP2 entire cores

```{r NGRIP2&GISP2, warning=FALSE, message=FALSE, fig.width=12, fig.height=6}
icecores %>%
  ggplot() + 
  geom_point(aes(x=years.BP, y=d18O.NGRIP2.ppt), color='blue', size=.5, alpha=.2) + 
  geom_point(aes(x=years.BP, y=d18O.GISP2.ppt), color='pink2', size=.5, alpha=.2) + 
  geom_smooth(aes(x=years.BP, y=d18O.NGRIP2.ppt, color='ngrip2', lty='ngrip2'), method='loess', span=.01, se=FALSE, lwd = 1)  + 
  geom_smooth(aes(x=years.BP, y=d18O.GISP2.ppt, color='gisp2', lty='gisp2'), method='loess', span=.01, se=FALSE, lwd = 1) + 
  scale_colour_manual(name=NULL, values =c('gisp2'='indianred3','ngrip2'='navy'), labels = c('GISP2','NGRIP2')) + 
  scale_linetype_manual(name=NULL, values =c('gisp2'='solid','ngrip2'='solid'), labels = c('GISP2','NGRIP2')) + 
  scale_x_reverse() + 
  #scale_x_continuous(breaks=c(19000,14000,10000), labels = c("19000","14000","10000"), trans = "reverse") + 
  geom_vline(xintercept = c(27000,22000,11000), lty='dashed', lwd=0.3) + 
  labs(x="years cal BP", 
       y="delta O18\ncolder << — >> warmer", 
       color="Cores",
       title="Paleoclimate Proxies from Greenland Ice Cores") +
  theme_minimal(base_size = 18) + 
  theme(legend.position="bottom", 
        plot.title = element_text(face = "bold", hjust = 0.5), 
        )
```

### Figure 3: Box Plots with Condensed Fitness Categories

```{r echo=T, fig.width=12, fig.height=12}
Hominin_demography %>% 
  mutate(fitness.aggregated = Fitness, 
         fitness.aggregated = recode(fitness.aggregated, 
                                     "birthrate=deathrate"="equal fitness", 
                                     "birthrate>deathrate"="equal fitness", 
                                     "MM higher mortality"="MM less fit", 
                                     "MM lower fertility"="MM less fit", 
                                     "MM lower mortality"="MM more fit", 
                                     "MN higher mortality"="MN less fit", 
                                     "MN lower fertility"="MN less fit", 
                                     "MN lower mortality"="MN more fit", 
                                     "NN higher mortality"="NN less fit", 
                                     "NN lower fertility"="NN less fit", 
                                     "NN lower mortality"="NN more fit"),
         fitness.aggregated = factor(fitness.aggregated, 
                                     levels = c(
                                       "equal fitness", 
                                       "MM less fit", 
                                       "MM more fit", 
                                       "MN less fit", 
                                       "MN more fit", 
                                       "NN less fit", 
                                       "NN more fit"))) %>% 
  ggplot(aes(x=phenotype, y=population)) + 
  geom_boxplot(fill="grey") + 
  facet_grid(fitness.aggregated~factor(home.range.radius), scales="free_y") + 
  labs(x="agent phenotype", y="population at end of simulation") + 
  theme_bw(base_size = 16)
```

### Figure 4: Example Model Trajectory with Logistical Foraging (Foraging Radius=12)

```{r echo=T, fig.width=10, fig.height=5}
Hominin_demography_trajectory %>% 
  ggplot(aes(y=population, x=step)) + 
  geom_line(aes(color=phenotype), size=2) +
  scale_color_manual(values = c("red", "orange", "grey", "forestgreen", "blue")) + 
  labs(x="model step", y="population", color="agent phenotype") + 
  theme_bw(base_size = 16) + 
  theme(legend.position="bottom")
```

## Demographic Dynamics of West Mediterranean

### Figure 5a: NGRIP and GISP for LGM to Holocene

```{r Late Pleistocene ice cores, echo=T, warning=FALSE, message=FALSE, fig.width=12, fig.height=5}
icecores %>% dplyr::filter(years.BP<=35000 & years.BP>=8000) %>% 
  ggplot() + 
  geom_line(aes(x=years.BP, y=d18O.GISP2.ppt), color='red', lwd=.2, alpha=.5) + 
  geom_line(aes(x=years.BP, y=d18O.NGRIP2.ppt), color='blue', lwd=.2, alpha=.5) + 
  geom_smooth(aes(x=years.BP, y=d18O.GISP2.ppt, color='gisp2', lty='gisp2'), method='loess', span=.07, se=FALSE, lwd = 1) + 
  geom_smooth(aes(x=years.BP, y=d18O.NGRIP2.ppt, color='ngrip2', lty='ngrip2'), method='loess', span=.07, se=FALSE, lwd = 1)  + 
  scale_colour_manual(name=NULL, values =c('gisp2'='red','ngrip2'='blue'), labels = c('GISP2','NGRIP2')) + 
  scale_linetype_manual(name=NULL, values =c('gisp2'='solid','ngrip2'='solid'), labels = c('GISP2','NGRIP2')) + 
  scale_x_reverse() + 
  #scale_x_continuous(breaks=c(19000,14000,10000), labels = c("19000","14000","10000"), trans = "reverse") + 
  #geom_vline(xintercept = c(19000,14000,10000), lty='dashed', lwd=0.3) + 
  labs(x="years cal BP", 
       y="delta O18\ncolder << — >> warmer", 
       color="Cores",
       title="Paleoclimate Proxies from Greenland Ice Cores") +
  theme_minimal(base_size = 18) + 
  theme(legend.position="none", 
        plot.title = element_text(face = "bold", hjust = 0.5), 
        )
```


### Figure 5b: Summed Probability Distribution (SPD) of Radiocarbon Dates in the West Mediterranean

#### Calibrate dates

```{r calibrate dates, message=FALSE, warning=FALSE, echo=T, results="hide"}
c14dates$BP.cal.median <- c14dates %>% 
  with(., calibrate(x=C14.mean, errors=C14.SD, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>% 
  medCal()
```


#### Calculate SPD

```{r SPD-all, message=FALSE, warning=FALSE, echo=FALSE, results="hide"}
date.bins <- c14dates %>% 
  dplyr::filter(calib.curve != "normal") %>%  
  with(., binPrep(site, C14.mean, 200))

model.all <- c14dates %>% 
  dplyr::filter(calib.curve != "normal") %>%  
  with(., calibrate(x=C14.mean, 
                    errors=C14.SD, 
                    calCurves = calib.curve, 
                    normalised=TRUE, 
                    calMatrix=FALSE)) %>% 
  modelTest(., timeRange = c(40000,7000), 
            errors = c14dates$C14.SD, 
            runm = 500, 
            model="exponential", 
            nsim = 200, 
            ncores = ncores,
            method = 'calsample', 
            bins = date.bins)
```


#### Graph SPD

```{r SPD-all graph, fig.height=4.55, fig.width=12, message=FALSE, warning=FALSE, echo=T, results="hide"}
par(mar=c(7,10,7,3))
plot(model.all, xlim = c(35000,8000), col.obs = 'red', lwd.obs = 3, drawaxes = F)
axis(1, cex.axis = 1, pos = -.0006)
axis(2, cex.axis = 1, pos = 35500)
mtext(side=1, line=3, "years cal BP", cex=1.3)
mtext(side=2, line=4, "summed probability", cex=1.3)
title(main=paste("West Mediterranean Late Pleistocene Assemblages (N = ", nrow(subset(c14dates, calib.curve != "normal")), ")"), cex.main = 1.5)
```



### Figure 5c: SPDs of Aggregated Technocomplexes

#### Classify aggregated technocomplexes

```{r classifiy technocomplexes, message=FALSE, warning=FALSE, echo=T, results="hide"}
c14dates <- c14dates %>% 
  mutate(industry.group = 
           case_when(
             str_detect(tolower(industry), "aurig") ~ "Aurignacian", 
             str_detect(tolower(industry), "gdrav") ~ "Gravettian",  
             str_detect(tolower(industry), "solut") ~ "Solutrean",
             str_detect(tolower(industry), "middle solutrean") ~ "Solutrean", 
             str_detect(tolower(industry), "salp") ~ "Solutrean",              
             str_detect(tolower(industry), "epig") ~ "Magdalenian",  
             str_detect(tolower(industry), "epi-gr") ~ "Magdalenian",
             str_detect(tolower(industry), "eppig") ~ "Magdalenian",
             str_detect(tolower(industry), "eppgr") ~ "Magdalenian",
             str_detect(tolower(industry), "apig") ~ "Magdalenian",
             str_detect(tolower(industry), "magd") ~ "Magdalenian",
             str_detect(tolower(industry), "badeg") ~ "Magdalenian",
             str_detect(tolower(industry), "gravet") ~ "Gravettian",  
             str_detect(tolower(industry), "gravt") ~ "Gravettian",  
             str_detect(tolower(industry), "azil") ~ "Epipaleolithic", 
             str_detect(tolower(industry), "romanell") ~ "Epipaleolithic", 
             str_detect(tolower(industry), "epipal") ~ "Epipaleolithic", 
             str_detect(tolower(industry), "mesol") ~ "Mesolithic", 
             str_detect(tolower(industry), "sauve") ~ "Mesolithic", 
             str_detect(tolower(industry), "tarden") ~ "Mesolithic", 
             str_detect(tolower(industry), "montcl") ~ "Mesolithic", 
             str_detect(tolower(industry), "montad") ~ "Mesolithic", 
             str_detect(tolower(industry), "castel") ~ "Mesolithic"
             ), 
         industry.group = factor(industry.group, 
                                 levels = c("Aurignacian", 
                                            "Gravettian", 
                                            "Solutrean", 
                                            "Magdalenian", 
                                            "Epipaleolithic", 
                                            "Mesolithic")),
         .after=industry)
```


#### Calculate SPDs of aggregated technocomplexes

```{r SPD-industries, message=FALSE, warning=FALSE, echo=T, results="hide"}
c14dates.industry.group <- c14dates %>% 
  dplyr::filter(!is.na(industry.group) & calib.curve != "normal")

c14dates.industry.group <- c14dates.industry.group[order(c14dates.industry.group$industry.group), ]

date.bins <-  c14dates.industry.group %>% 
  with(., binPrep(site, C14.mean, 200))

stacked.spd <- c14dates.industry.group %>% 
  with(., calibrate(x=C14.mean, errors=C14.SD, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>% 
  stackspd(timeRange=c(40000,7000), 
           bins=date.bins,
           group=c14dates.industry.group$industry.group, 
           runm = 500, 
           datenormalised=TRUE)
```


#### Graph SPDs

```{r SPD-industries graph, fig.width=12, fig.height=6, echo=T}
par(mar=c(7,10,7,3))
options(digits = 7)
plot(stacked.spd, 
     xlim = c(35000,7500), 
     rescale = T, 
     spdnormalised = F, 
     type = "multipanel", 
     lwd.obs = .5, 
     runm = NA, 
     cex.lab = 1.3, 
     cex.axis = 1, 
     ylab = "summed probability\n", 
     xlab = "years cal BP",
     ymargin = 1, 
     gapFactor = .4, 
     legend.arg=list("topleft", bty = "n", horiz = T, cex = 1))
#axis(1, cex.axis = 2, pos = -.0006)
#axis(2, cex.axis = 2, pos = 35500)
#mtext(side=1, line=5, "years cal BP", cex=2.5)
#mtext(side=2, line=6, "summed probability", cex=2.5)
title(main=paste("West Mediterranean Late Pleistocene Technocomplexes (N = ", nrow(c14dates.industry.group), ")"), cex.main = 1.5)
```
#### Permutation test of aggregated technocomplexes

```{r SPD-perm-industries, message=FALSE, warning=FALSE, echo=FALSE, results="hide"}
c14dates.industry.group <- c14dates %>% 
  dplyr::filter(!is.na(industry.group) & calib.curve != "normal")

c14dates.industry.group <- c14dates.industry.group[order(c14dates.industry.group$industry.group), ]

date.bins <-  c14dates.industry.group %>% 
  with(., binPrep(site, C14.mean, 200))

perm.test <- c14dates.industry.group %>% 
  with(., calibrate(x=C14.mean, errors=C14.SD, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>% 
  permTest(timeRange=c(40000,7000), 
           bins=date.bins,
           marks=c14dates.industry.group$industry.group, 
           nsim = 200, 
           runm = 500, 
           datenormalised=TRUE)
```


#### Graph Permutations

```{r SPD-perm-industries graph, fig.width=12, fig.height=3, echo=T}
plot(perm.test, 
     focalm = "Aurignacian", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.04),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Aurignacian (N = ", nrow(subset(c14dates, industry.group == "Aurignacian")), ")"), cex.main = 1.5)
 
plot(perm.test, 
     focalm = "Gravettian", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.04),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Gravettian (N = ", nrow(subset(c14dates, industry.group == "Gravettian")), ")"), cex.main = 1.5)
  
plot(perm.test, 
     focalm = "Solutrean", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.04),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Solutrean (N = ", nrow(subset(c14dates, industry.group == "Solutrean")), ")"), cex.main = 1.5)
  
plot(perm.test, 
     focalm = "Magdalenian", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.1),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Magdalenian (N = ", nrow(subset(c14dates, industry.group == "Magdalenian")), ")"), cex.main = 1.5)
  
plot(perm.test, 
     focalm = "Epipaleolithic", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.04),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Epipaleolithic (N = ", nrow(subset(c14dates, industry.group == "Epipaleolithic")), ")"), cex.main = 1.5)
  
plot(perm.test, 
     focalm = "Mesolithic", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.12), 
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Mesolithic (N = ", nrow(subset(c14dates, industry.group == "Mesolithic")), ")"), cex.main = 1.5)
```

#### Permutation by region

```{r SPD-perm-region, message=FALSE, warning=FALSE, echo=FALSE, results="hide"}
date.bins <-  c14dates %>% 
  with(., binPrep(site, C14.mean, 200))

perm.test <- c14dates %>% 
  with(., calibrate(x=C14.mean, errors=C14.SD, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>% 
  permTest(timeRange=c(40000,7000), 
           bins=date.bins,
           marks=c14dates$region, 
           nsim = 200, 
           runm = 500, 
           datenormalised=TRUE)
```

#### Graph Permutations

```{r SPD-perm-region graph, fig.width=12, fig.height=4, echo=T}
plot(perm.test, 
     focalm = "Southwest", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.04),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Southwest (N = ", nrow(subset(c14dates, region == "Southwest")), ")"), cex.main = 1.5)
 
plot(perm.test, 
     focalm = "N Central", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.1),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("North Central (N = ", nrow(subset(c14dates, region == "N Central")), ")"), cex.main = 1.5)
  
plot(perm.test, 
     focalm = "Northeast", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.04),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Northeast (N = ", nrow(subset(c14dates, region == "Northeast")), ")"), cex.main = 1.5)
  
plot(perm.test, 
     focalm = "Southeast", 
     xlim = c(35000,8000), 
     col.obs = 'red', 
     lwd.obs = 3,
     ylim = c(0, 0.04),
     drawaxes = F)
  axis(1, cex.axis = 1, pos = -.0006)
  axis(2, cex.axis = 1, pos = 35500)
  mtext(side=1, line=3, "years cal BP", cex=1.3)
  mtext(side=2, line=4, "summed probability", cex=1.3)
  title(main=paste("Southeast (N = ", nrow(subset(c14dates, region == "Southeast")), ")"), cex.main = 1.5)
  
```